---
title: "Xcertainty"
output: 
  html_document:
    keep_md: yes
description: >
  Start here to learn how to use Xcertainty. You'll learn how to include drone-based measurement data into a Bayesian statistical model to produce predictive posterior distributions that can be used to describe each measurement and its associated uncertainty.
vignette: >
  %\VignetteIndexEntry{Xcertainty}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
fontsize: 13pt 
author: "KC Bierlich & Josh Hewitt, CODEX"
date: "2024-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "png", cache = TRUE)
```


## Introduction
All morphological measurements derived using drone-based photogrammetry are susceptible to uncertainty. This uncertainty often varies by the drone system used. Thus, it is critical to incorporate photogrammetric uncertainty associated with measurements collected using different drones so that results are robust and comparable across studies and over long-term datasets.  

The `Xcertainty` package makes this simple and easy by producing a predictive posterior distribution for each measurement. This posterior distribution can be summarized to describe the measurement (i.e., mean, median) and its associated uncertainty (i.e., standard deviation, credible intervals). The posterior distributions are also useful for making probabilistic statements, such as classifying maturity or diagnosing pregnancy if a proportion of the posterior distribution for a given measurement is greater than a specified threshold (e.g., if greater than 50% of posterior distribution for total body length is > 10 m, the individual is classified as mature).

`Xcertainty` is based off the Bayesian statistical model described in [Bierlich et al., 2021a](https://doi.org/10.3354/meps13814) where measurements of known-sized objects ('calibration objects') collected at various altitudes are used as training data to predict morphological measurements (e.g., body length) and associated uncertainty of unknown-sized objects (e.g., whales). This modeling approach was later adapted to incorporate multiple measurements (body length and width) to estimate body condition with associated uncertainty  [Bierlich et al. (2021b)](https://doi.org/10.3389/fmars.2021.749943), as well as combine body length with age information to construct growth curves [Bierlich et al., 2023](https://doi.org/10.1098/rsbl.2023.0043) and [Pirotta and Bierlich et al., 2024](https://doi.org/10.1111/gcb.17366).

In this vignette, we'll cover how to setup your data, run `Xcertainty`, calculate body condition metrics, and interpret results.  



## Main inputs

`Xcertainty` follows these main steps.      

1. Prepare calibration and observation data:  
    + `parse_observations()`: parses wide-format data into a normalized list of dataframe objects. 

&nbsp;

2. Build sampler:   
    + 2a. Calibration Objects
      + `calibration_sampler()`: estimate measurement error parameters for calibration/training data.   
    + 2b. Observation Data
      + `independent_length_sampler()`: this model assumes all Subject/Measurement/Timepoint combinations are independent. So this is well suited for data that contains individuals that either have no replicate samples or have replicate samples that are independent over time, such as body condition which can increase or decrease, as opposed to length which should be stable or increase over time.
  
      + `nondecreasing_length_sampler()`: data contains individuals with replicate samples for length over time but no age information. This sampler sets a rule so that length measurements of an individual cannot shrink over time (from year to year), i.e., an individual should not (in most cases!) be getting shorter over time. 
  
      + `growth_curve_sampler()`: data contains individuals with replicate samples and age information. This model fits a Von Bertalanffy-Putter growth curve to observations following [Pirotta and Bierlich et al., 2024](https://doi.org/10.1111/gcb.17366). 

&nbsp;

3. Run!  
    + `sampler()`: this function runs the sampler that you built. You can set the number of iterations using 'niter'.


## Example: Gray whale body length and body condition
We will use a small example dataset consisting of body length and body width measurements of Pacific Coast Feeding Group (PCFG) gray whales imaged along the coast of Newport, Oregon, USA using two different drones, a DJI Inspire 2 (I2, n = 5 individuals) and a DJI Phantom 4 Pro (P4P, n = 5 individuals). The P4P contains only a barometer for estimating altitude, while the I2 contains both a barometer and a LiDAR (or laser) altimeter (LidarBoX, [Bierlich et al., 2024](https://doi.org/10.1139/dsa-2023-0051)). In this example, we will use data from the I2 (see [Xcertianty_informative_priors](Xcertainty_informative_priors.html) for an example using P4P data).

We will use the length and width measurements to calculate several body condition metrics (body area index, body volume, surface area, and 
standardized widths). We used open-source software [MorphoMetriX v2](https://github.com/MMI-CODEX/MorphoMetriX-V2) and [CollatriX](https://github.com/cbirdferrer/collatrix) to measure the whales and process the measurements, respectively.


Steps:     
1. Prepare calibration data and observation (whale) data.    
2. Build sampler.   
3. Run the sampler.   
4. Calculate body condition metrics.   
5. View outputs.    


&nbsp;

We'll first load the Xcertainty package, as well as other packages we will use throughout this example.
```{r, echo=TRUE, warning=FALSE}
library(Xcertainty)

library(tidyverse)
library(ggdist)
```

### Calibration Objects
First we'll load and prepare the calibration data, which is from [Bierlich et al., 2024](https://doi.org/10.1139/dsa-2023-0051). Note that "CO" here stands for "Calibration Object" used for training data, and "CO.L" is the true length of the CO (1 m) and "Lpix" is the photogrammetric measurement of the CO in pixels. Each UAS has a unique CO.ID so that the training data and observation (whale) data can be linked. We will filter to use CO data from the I2 drone.
```{r}
# load calibration measurement data
data("co_data")

# sample size for both drones
table(co_data$uas)

# filter for I2 drone
co_data_I2 <- co_data %>% filter(uas == "I2")
```


Next, well format the data using `parse_observations()`.
```{r}
calibration_data = parse_observations(
  x = co_data_I2, 
  subject_col = 'CO.ID',
  meas_col = 'Lpix', 
  tlen_col = 'CO.L', 
  image_col = 'image', 
  barometer_col = 'Baro_Alt',
  laser_col = 'Laser_Alt', 
  flen_col = 'Focal_Length', 
  iwidth_col = 'Iw', 
  swidth_col = 'Sw',
  uas_col = 'uas'
)
```

This creates a list of four elements:   
    * `calibration_data$pixel_counts`.   
    * `calibration_data$training_objects`.    
    * `calibration_data$image_info`.
    * `calibration_data$prediction_objects`
 

### Gray whale measurements
Now we'll load and prepare the gray whale measurement data. The column 'whale_ID' denotes the unique individual. Note, some individuals have multiple images --  Xcertainty incorporates measurements across images for an individual to produce a single posterior distribution for the measurement of that individual. For example, multiple body length measurements from different images of an individual will produce a single posterior distribution of body length for that individual. 

To estimate body condition for these gray whales, we will use body widths between 20-70% of the body length. We'll save the column names of these widths as their own object.  

For this example, we will only use whale measurements collected using the I2 drone. See the vignette titled "Xcertainty_informative_prios" for an example using P4P data. Note, that Xcertainty can also incorporate measurements with missing LiDAR data (NAs).
```{r}
# load gray whale measurement data
data("gw_data")

# quick look at the data
head(gw_data)

# number of images per whale ID
table(gw_data$whale_ID)


# filter for I2 drone and select specific widths to include for estimating body condition (20-70%)
gw_measurements <- gw_data %>% filter(uas == "I2") %>% 
  select(!c("TL_w05.00_px", "TL_w10.00_px", "TL_w15.00_px", 
            "TL_w75.00_px", "TL_w80.00_px", "TL_w85.00_px", "TL_w90.00_px", "TL_w95.00_px"))

# identify the width columns in the dataset
width_names = grep(
  pattern = 'TL_w\\_*', 
  x = colnames(gw_measurements),
  value = TRUE
)
```

&nbsp;

Next, we'll use `parse_observations()` to prepare the whale data. Since `Xcertainty` incorporates errors associated with both a LiDAR altimeter and a barometer into the output measurement, the input measurements must be in pixels. In our example dataset of gray whales, measurements are already in pixels. If measurements in a dataframe are in meters, they can easily be converted into pixels using `alt_conversion_col` to assign which altitude column should be used to "back calculate" measurements in meters into pixels. For example, use `alt_conversion_col = 'Baro_Alt` if the measurements used the barometer to convert measurements into meters.   

Note that you can also set the specific timepoint to link measurements of individuals using `timepoint_col`. For example, if you wanted all the total body length measurements of an individual included to produce a single length measurement over the course of the season, you may choose `timepoint_col = 'year'`, or you may want body condition at the daily level, so you could enter `timepoint_col = 'date'`. In our example, measurements are already synced at the daily level, so we will keep default as is.

&nbsp;

Also, note that we assign the measurement column (`meas_col`) for TL and the widths between 20-70% that we saved as "width_names". 
```{r}
# parse field study
whale_data = parse_observations(
  x = gw_measurements, 
  subject_col = 'whale_ID',
  meas_col = c('TL_px', width_names),
  image_col = 'image', 
  barometer_col = 'Baro_Alt',
  laser_col = 'Laser_Alt', 
  flen_col = 'Focal_Length', 
  iwidth_col = 'Iw', 
  swidth_col = 'Sw', 
  uas_col = 'uas'
  #alt_conversion_col = 'altitude'
)
```

  
This creates a list of four elements:   
    * `calibration_data$pixel_counts`.   
    * `calibration_data$training_objects`.    
    * `calibration_data$image_info`.
    * `calibration_data$prediction_objects`


## Build sampler
Now we will build a sampler. Always start with using non-informative priors, which should be appropriate for most datasets. In some cases, using informative priors may be more appropriate, especially for datasets that have high errors and when the model is overparameterized -- see the vignette "Xcertianty_informative_priors" for an example. We'll set the altitudes (`image_altitude`) and object length measurements (`object_lengths`) to cover an overly wide range for our target species.  
```{r}
sampler = independent_length_sampler(
  data = combine_observations(calibration_data, whale_data),
  priors = list(
    image_altitude = c(min = 0.1, max = 130),
    altimeter_bias = rbind(
      data.frame(altimeter = 'Barometer', mean = 0, sd = 1e2),
      data.frame(altimeter = 'Laser', mean = 0, sd = 1e2)
    ),
    altimeter_variance = rbind(
      data.frame(altimeter = 'Barometer', shape = .01, rate = .01),
      data.frame(altimeter = 'Laser', shape = .01, rate = .01)
    ),
    altimeter_scaling = rbind(
      data.frame(altimeter = 'Barometer', mean = 1, sd = 1e1),
      data.frame(altimeter = 'Laser', mean = 1, sd = 1e1)
    ),
    pixel_variance = c(shape = .01, rate = .01),
    object_lengths = c(min = .01, max = 20)
  )
)
```


## Run Sampler
Now we can run the sampler. Note, that "niter" refers to the number of iterations. When exploring data outputs, 1e4 or 1e5 can be good place for exploration, as this won't take too much time to run. We recommend using 1e6 for the final analysis since 1e6 MCMC samples is often enough to get a reasonable posterior effective sample size. 
```{r}
# run sampler
output = sampler(niter = 1e6, thin = 10)
```



## View Sampler Outputs (TL and widths) 
Our saved `output` contains all the posterior samples and summaries of all training data and length and width measurements from the sampler. **Note**, that there are many objects stored in `output`, so it is best to view specific selections rather than viewing all of the objects stored in `output` at once, as this can take a very long time to load and cause R to freeze. 

&nbsp;

We can view the posterior summaries (mean, sd, etc.) for each altimeter. Note that the `lower` and `upper` represent the 95% highest posterior density intervals (HPDI) of the posterior distribution.
```{r}
output$summaries$altimeters
```

&nbsp;

And we can view and compare the posterior outputs for each image's altitude compared to the observed altitude from the barometer (blue) and LiDAR (orange) in the training dataset.
```{r, Xc_Fig1_posterior_vs_obs_alt, fig.dim = c(7, 5)}
output$summaries$images %>% left_join(co_data %>% rename(Image = image), by = "Image") %>%
  ggplot() + theme_bw() + 
  geom_pointrange(aes(x = Baro_Alt, y = mean, ymin = lower, ymax = upper), color = "blue") + 
  geom_pointrange(aes(x = Laser_Alt, y = mean, ymin = lower, ymax = upper), color = "orange") + 
  geom_abline(slope = 1, intercept = 0, lty = 2) + 
  ylab("posterior altitude (m)") + xlab("observed altitude (m)")

```

&nbsp;

We can also view the pixel variance from the training data
```{r}
output$pixel_error$summary
```


&nbsp;

We can view the posterior summaries (mean, sd, etc.) for all measurements of each individual whale. As above, the `lower` and `upper` represent the 95% HPDI of the posterior distribution for that specific measurement.
```{r}
head(output$summaries$objects)
```

&nbsp;

You can filter to view a specific measurement across all individuals, such as total body length (TL). 
```{r}
output$summaries$objects %>% filter(Measurement == "TL_px")
```

&nbsp;

Or filter directly for all measurements from a specific individual
```{r}
output$summaries$objects %>% filter(Subject == "GW_01")
```

&nbsp;

Plot total body length with associated uncertainty for each individual. Here the black dots represent the mean of the posterior distribution for total body length and the black bars around each dot represents the uncertainty, as 95% HPDI.  
```{r, Xc_Fig2_TL_per_subject, fig.dim = c(7, 5)}
output$summaries$objects %>% filter(Measurement == "TL_px") %>% 
  ggplot() + theme_bw() + 
  geom_pointrange(aes(x = Subject, y = mean, ymin =lower, ymax = upper)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) + 
  ylab("Total body length (m)") 
```

&nbsp;

You can also view and plot the posterior samples for an individual's measurement. Note, be sure to exclude the first half of the posterior samples, i.e., if 10000 samples, exclude the first 5000. To demonstrate this below, we'll first save the samples for an individual as an object, then plot the distribution with the first half of the samples excluded. 
```{r, Xc_Fig3_TL_dist, fig.dim = c(6, 6)}
ID_samples <- output$objects$`GW_01 TL_px 1`$samples

data.frame(TL = ID_samples[(length(ID_samples)/2):length((ID_samples))]) %>%
  ggplot() + stat_halfeye(aes(TL), .width = 0.95) + theme_bw() 
```


&nbsp;


### Body Condition 

We'll also calculate body condition from the posterior samples using `body_condition()`, which calculates several body condition metrics using the posterior distributions of the body widths:  

  + body area index (BAI) [Burnett et al., 2018](https://doi.org/10.1111/mms.12527); [Bierlich et al. (2021b)](https://doi.org/10.3389/fmars.2021.749943)  
  
  + body volume [Christiansen et al. 2021](https://doi.org/10.3354/meps13585)  
  
  + surface area [Christiansen et al., 2016](https://doi.org/10.1002/ecs2.1468)  
  
  + standardized body widths (standardized by TL).   


Note, for calculating body volume, the default for `height_ratios` is 1, implying that the vertical cross section of the animal is circular rather than elliptical. To calculate body volume using an elliptical model [i.e., Christiansen et al., 2019](https://doi.org/10.1111/2041-210X.13298), enter the the height-width ratio for each width using `height_ratios`. 
 
```{r}
# First, enumerate the width locations along the animal's length
width_increments = as.numeric(
  str_extract(
    string = width_names, 
    pattern = '[0-9]+'
  )
)

# Compute body condition
body_condition_output = body_condition(
  data = whale_data, 
  output = output,
  length_name = 'TL_px',
  width_names = width_names,
  width_increments = width_increments,
  summary.burn = .5,
  height_ratios = rep(1, length(width_names)) # assumes circular cross section
)
```

&nbsp; 

Note, there are a lot of objects stored in the `body_condition_output`, so it's best to view selected outputs rather than all objects at once, as it may take a long time to load and can freeze R.

You can view the body condition summaries (`surface_area`, `body_area_index`, `body_volume`, and `standardized_widths`) across individuals using `body_condition_output$summaries`. Summaries include mean, standard deviation (sd) and the lower and upper 95% HPDI. 

&nbsp;

View summary of BAI
```{r}
head(body_condition_output$summaries$body_area_index)
```

&nbsp;

Plot BAI results 
```{r, Xc_Fig4_bai, fig.dim = c(7, 5)}
body_condition_output$summaries$body_area_index %>% 
  ggplot() + theme_bw() + 
  geom_pointrange(aes(x = Subject, y = mean, ymin =lower, ymax = upper)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) + 
  ylab("Body Area Index") 
```


&nbsp;


View summary of Body Volume
```{r}
head(body_condition_output$summaries$body_volume)
```

&nbsp;

Plot Body Volume results 
```{r, Xc_Fig5_body_vol, fig.dim = c(7, 5)}
body_condition_output$summaries$body_volume %>% 
  ggplot() + theme_bw() + 
  geom_pointrange(aes(x = Subject, y = mean, ymin =lower, ymax = upper)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) + 
  ylab("Body Volume (m^3)") 
```

&nbsp;

View the standardized widths of an individual
```{r}
body_condition_output$summaries$standardized_widths %>% filter(Subject == "GW_01")
```

&nbsp;

Plot standardized widths of an individual
```{r, Xc_Fig6_std_widths, fig.dim = c(7, 5)}
body_condition_output$summaries$standardized_widths$metric <- gsub("standardized_widths TL_", "", body_condition_output$summaries$standardized_widths$metric)

body_condition_output$summaries$standardized_widths %>% filter(Subject == "GW_01") %>%
  ggplot() + theme_bw() + 
  geom_pointrange(aes(x = metric, y = mean, ymin = lower, ymax = upper)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("width%") + ylab("standardized width") + ggtitle("GW_01") 
```


&nbsp;

Can also view standardized widths across all individuals
```{r, Xc_Fig7_std_widths_all, fig.dim = c(7, 5)}
body_condition_output$summaries$standardized_widths %>% 
  ggplot() + theme_bw() + 
  geom_boxplot(aes(x = metric, y = mean)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("width%") + ylab("standardized width")
```

&nbsp;

You can also view individual's posterior samples for any of the body condition metrics.
```{r}
head(body_condition_output$body_area_index$`GW_01`$samples)
```



And from these posterior samples for an individual, we can plot the distribution. Here we'll plot BAI as an example, and include the 95% HPDI. Remember to exclude the first half of the samples, as mentioned above. 
```{r, Xc_Fig8_bai_dist, fig.dim = c(6, 6)}
ID_samples <- body_condition_output$body_area_index$`GW_01 1`$samples
  
data.frame(BAI = ID_samples[(length(ID_samples)/2):length((ID_samples))]) %>%
  ggplot() + stat_halfeye(aes(BAI), .width = 0.95) + theme_bw() + ggtitle("GW_01")
```


&nbsp;

We hope this vignette has been helpful for getting started with organizing your input data and how to view and interpret results. Check out our other vignettes to view examples of other samplers, including using "informative priors" 
and "growth curves". 


